Dissecting the causal question underlying the association between cancer and ADRD 

### 2021-03-24 04_reports/manuscript.Rmd
update on competing events

Death is a competing event of ADRD because if a participant dies prior to ADRD diagnosis, death prevents from observing ADRD at future time-points. The causal diagram in Figure 4 is an extension of Figure 2, including $D_{80}$ as a representation of Death by age 80. We include an arrow between $P$ and $D_{80}$ since $P$ may increase the risk, through tumor development or other shared common causes. Furthermore we include an arrow between $P^*$ and $D_{80}$ since cancer diagnosis, and subsequent treatment (or lack of treatment) may have an effect on death. The arrow from $D_{80}$ and $Y_{85}$ represents the key feature of a competing events data structure: an individual who dies by age 80 cannot subsequently develop ADRD. Last, since $D_{80}$ and $Y_{85}$ events that are related to the aging process, $C$ represent the shared causes related to both events (sch as cardiovascular conditions, fraility, other comorbidities, etc).

If we had measured all $L$, we could estimate the total effect of $P$ in the risk of ADRD under no additional assumptions. This effect includes all the pathways including those mediaded through death.

There are several causal estimands that can help us identify the direct of $P$ in $Y_{85}$ as presented by Young et al. 
However if we want to isolate the direct effect of P in Y (path in violet), we

>>>>

As illustrated in previous figures, we are interested in the direct effect of Pin1 in ADRD at 85 years, which means a length of follow-up of 20 years. However, mortality in elderly population between 65 and 85 is substancial, and a leading cause of death is cancer. This means that in order to measure ADRD by the age of 85, individuals have to had survive to cancer and to other causes. For this reason death is a competing event of ADRD because if a participant dies prior to ADRD diagnosis, death prevents from observing ADRD at future time-points.

We can visualize this data feature in the causal diagram in Figure 4. In this DAG we include $D_{80}$ as a representation of Death by age 80, an arrow between $P$ and $D_{80}$ since $P$ may increase the mortality risk irrespective of cancer diagnosis. We also include an arrow between $P^*$ and $D_{80}$ since cancer diagnosis, and subsequent treatment (or lack of treatment) may have an effect on death. Last, the arrow from $D_{80}$ and $Y_{85}$ represents the key feature of a competing events data structure: an individual who dies by age 80 cannot subsequently develop ADRD, and since $D_{80}$ and $Y_{85}$ are events related to aging, $C$ represent the shared causes of both events.

In this DAG we observe that, had we measured and adjusted for $L$, we could estimate the total effect of $P$ in $Y_{85}$ without further assumptions. However the total effect is a causal estimand that includes both the direct pathway between $P$ and $Y_{85}$. This means that if Pin1 has an effect on mortality through cancer or other mechanisms, we will observe a protective effect of Pin1 overexpression in ADRD, partially mediated through death.

If the aim is to isolate the direct effect of $P$ in $Y_{85}$ as in Figure 4 (violet arrows), we first need to define the causal estimand of interest. There are at least three causal estimands, each one has different interpretation and relies on different assumptions: the controlled direct effect, the survivor average causal effect and the separable direct effect. Through the next section we will discuss the controlled direct effect since this estimand is tied to frequently used estimators.

There are several ways to conceptualize direct effects and there are several causal parameters (estimands) to represent it, such as: the controlled direct effect (CDE), the survivor average causal effect and the separable direct effect. Through this section we will discuss the causal question, assumptions related to death and interpretation of the CDE since this estimand is connects to frequently used estimators.

THe CDE represents the effect of Pin1 in ADRD in a setting where we could have prevented death over the study period.
